Longitudinal Principal Manifold Estimation

### 2023-08-04 paper/main.tex
update application section and add discussion

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.223 (0.311) & 0.524 (0.462) & N/A & 25.5 (75.3) & 0.984 (0.754) & 2.00 (9.29) & N/A & N/A & N/A \\
    LPME & 0.138 (0.174) & 0.363 (0.342) & N/A & 19.8 (69.1) & 0.751 (0.660) & 1.83 (9.19) & N/A & N/A & N/A \\
    PME & 0.231 (0.412) & 0.871 (1.76) & N/A & 25.5 (75.3) & 1.11 (1.36) & 2.02 (9.29) & N/A & N/A & N/A \\
    PC/PS & 0.189 (0.296) & 0.615 (0.357) & N/A & 25.5 (75.3) & 0.928 (0.702) & 1.25 (3.69) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.124 (0.187) & 0.385 (0.640) & N/A & 4.11 (15.9) & 0.939 (1.26) & 0.272 (0.926) & N/A & N/A & N/A \\
    LPME & 0.0707 (0.138) & 0.251 (0.511) & N/A & 3.23 (7.67) & 0.611 (1.05) & 0.290 (0.860) & N/A & N/A & N/A \\
    PME & 0.105 (0.202) & 0.379 (0.708) & N/A & 4.06 (15.7) & 0.933 (1.32) & 0.295 (0.938) & N/A & N/A & N/A \\
    PC/PS & 0.101 (0.177) & 0.504 (0.413) & N/A & 4.06 (15.9) & 0.830 (1.09) & 0.556 (0.317) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Median (IQR).}
  \label{table:simulation_results_median}
\end{table}

@@

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.

Provide a brief summary of the results of the model fitting here, along with the related figures.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{adni_lhipp}
  \caption{Left Hippocampus, ADNI Participant}
  \label{fig:adni_result}
\end{figure}

@@

\subsection*{Contributions}

\subsection*{Limitations}

\begin{enumerate}
  \item Algorithm assumes stability in shape over time.
  \item Algorithm initialization is sensitive to magnitude of change between time points.
  \item Computation time.
\end{enumerate}

Future Work:

>>>>

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.222 (0.261) & 0.513 (0.411) & N/A & 27.3 (74.9) & 0.986 (0.787) & 1.51 (6.04) & N/A & N/A & N/A \\
    LPME & 0.129 (0.161) & 0.417 (0.960) & N/A & 19.0 (63.0) & 0.776 (0.777) & 1.27 (5.71) & N/A & N/A & N/A \\
    PME & 0.247 (0.435) & 0.854 (1.62) & N/A & 27.3 (74.9) & 1.04 (1.07) & 1.55 (6.04) & N/A & N/A & N/A \\
    PC/PS & 0.188 (0.247) & 0.600 (0.303) & N/A & 27.3 (74.9) & 0.939 (0.730) & 1.04 (2.20) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.144 (0.222) & 0.466 (0.658) & N/A & 4.17 (19.4) & 0.890 (1.39) & 0.305 (1.07) & N/A & N/A & N/A \\
    LPME & 0.0727 (0.129) & 0.250 (0.477) & N/A & 3.23 (11.3) & 0.575 (1.08) & 0.293 (0.907) & N/A & N/A & N/A \\
    PME & 0.123 (0.249) & 0.516 (0.757) & N/A & 4.13 (19.2) & 0.897 (1.39) & 0.331 (1.04) & N/A & N/A & N/A \\
    PC/PS & 0.112 (0.205) & 0.556 (0.427) & N/A & 4.12 (19.3) & 0.805 (1.26) & 0.567 (0.388) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Median (IQR).}
  \label{table:simulation_results_median}
\end{table}

@@

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin. Finally, the data were augmented with spherical coordinates in a manner similar to that described in the previous section to avoid the need to fit to a self-intersecting manifold.

\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}. Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation. The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.

Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots. Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{adni_lhipp}
  \caption{Left Hippocampus, Cognitively Healthy ADNI Participant}
  \label{fig:adni_result}
\end{figure}

\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{adni_lhipp_cross_section}
  \caption{Left Hippocampus Cross Section}
  \label{fig:adni_cross_section}
\end{figure}

@@

In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data. To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a longitudinal setting. In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point. This improvement was seen across several manifold types. We also suggest a data augmentation approach that may allow for the circumvention of the inability to fit principal curve-based methods to some self-intersecting manifolds.

While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations. First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly. Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that are too curvy. The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern. There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.

A second limitation is that the LPME method requires relative stability in the overall shape of the structure being fit over time. Thus, while this algorithm can be useful for separating meaningful changes in a manifold over time from inter-observation error, it will be unable to fit in situations where the error between observations is too substantial. How restrictive this requirement is depends largely on the area of application. When used to estimate subcortical surfaces as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. This shortcoming could potentially be addressed in several ways. First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section 4.

Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities. The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin. The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations. While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.

\newpage
